---
title: "Is it time for an NBA expansion?"
subtitle: "My subtitle if needed"
author: 
  - Yan Mezhiborsky
thanks: "Code and data are available at: ."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(readr)

```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}


NBA1980 <- read_csv("../data/raw_data/NBA1980.csv")
NBA1981 <- read_csv("../data/raw_data/NBA1981.csv")
NBA1982 <- read_csv("../data/raw_data/NBA1982.csv")
NBA1983 <- read_csv("../data/raw_data/NBA1983.csv")
NBA1984 <- read_csv("../data/raw_data/NBA1984.csv")
NBA1985 <- read_csv("../data/raw_data/NBA1985.csv")
NBA1986 <- read_csv("../data/raw_data/NBA1986.csv")
NBA1987 <- read_csv("../data/raw_data/NBA1987.csv")
NBA1988 <- read_csv("../data/raw_data/NBA1988.csv")
NBA1989 <- read_csv("../data/raw_data/NBA1989.csv")
NBA1990 <- read_csv("../data/raw_data/NBA1990.csv")
NBA1991 <- read_csv("../data/raw_data/NBA1991.csv")
NBA1992 <- read_csv("../data/raw_data/NBA1992.csv")
NBA1993 <- read_csv("../data/raw_data/NBA1993.csv")
NBA1994 <- read_csv("../data/raw_data/NBA1994.csv")
NBA1995 <- read_csv("../data/raw_data/NBA1995.csv")
NBA1996 <- read_csv("../data/raw_data/NBA1996.csv")
NBA1997 <- read_csv("../data/raw_data/NBA1997.csv")
NBA1998 <- read_csv("../data/raw_data/NBA1998.csv")
NBA1999 <- read_csv("../data/raw_data/NBA1999.csv")
NBA2000 <- read_csv("../data/raw_data/NBA2000.csv")
NBA2001 <- read_csv("../data/raw_data/NBA2001.csv")
NBA2002 <- read_csv("../data/raw_data/NBA2002.csv")
NBA2003 <- read_csv("../data/raw_data/NBA2003.csv")
NBA2004 <- read_csv("../data/raw_data/NBA2004.csv")
NBA2005 <- read_csv("../data/raw_data/NBA2005.csv")
NBA2006 <- read_csv("../data/raw_data/NBA2006.csv")
NBA2007 <- read_csv("../data/raw_data/NBA2007.csv")
NBA2008 <- read_csv("../data/raw_data/NBA2008.csv")
NBA2009 <- read_csv("../data/raw_data/NBA2009.csv")
NBA2010 <- read_csv("../data/raw_data/NBA2010.csv")
NBA2011 <- read_csv("../data/raw_data/NBA2011.csv")
NBA2012 <- read_csv("../data/raw_data/NBA2012.csv")
NBA2013 <- read_csv("../data/raw_data/NBA2013.csv")
NBA2014 <- read_csv("../data/raw_data/NBA2014.csv")
NBA2015 <- read_csv("../data/raw_data/NBA2015.csv")
NBA2016 <- read_csv("../data/raw_data/NBA2016.csv")
NBA2017 <- read_csv("../data/raw_data/NBA2017.csv")
NBA2018 <- read_csv("../data/raw_data/NBA2018.csv")
NBA2019 <- read_csv("../data/raw_data/NBA2019.csv")
NBA2020 <- read_csv("../data/raw_data/NBA2020.csv")
NBA2021 <- read_csv("../data/raw_data/NBA2021.csv")
NBA2022 <- read_csv("../data/raw_data/NBA2022.csv")
NBA2023 <- read_csv("../data/raw_data/NBA2023.csv")

```

```{r}

#combining data into one table
combined_data <- data.frame()

# Loop through the years 1980 to 2023
for(year in 1980:2023) {
  # Read the CSV file for the year
  file_path <- sprintf("../data/raw_data/NBA%d.csv", year)
  yearly_data <- read_csv(file_path)
  
  # Add a year column
  yearly_data$Year <- year
  
  # Combine with the main data frame
  combined_data <- bind_rows(combined_data, yearly_data)
}

# creating table for league average
league_average_data <- combined_data %>% 
  filter(Team == "League Average")

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it. 



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```






# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References


